---
title: "RNA-seq Data Analysis Assignment"
output: html_document
date: "2025-09-20"
---

## Introduction
This analysis explores RNA-seq differential expression data from an experiment comparing a diseased cell line with the same cell line treated with compound X. The main objectives are to:

- Generate a volcano plot to visualize differential expression.
- Identify significantly upregulated and downregulated genes.
- Report the top 5 upregulated and downregulated genes.
- Briefly describe the biological functions of these genes.
`
```{r}
# Load data directly from the link
url <- "https://gist.githubusercontent.com/stephenturner/806e31fce55a8b7175af/raw/1a507c4c3f9f1baaa3a69187223ff3d3050628d4/results.txt"
rna_data <- read.table(url, header = TRUE, sep = "", stringsAsFactors = FALSE)

```

```{r}
# Inspect data
head(rna_data)
str(rna_data)
colnames(rna_data)
```

```{r}
# Classify as Upregulated, Downregulated, or Not Significant
rna_data$Significant <- ifelse(
  rna_data$log2FoldChange > 1 & rna_data$pvalue < 0.01, "Upregulated",
  ifelse(rna_data$log2FoldChange < -1 & rna_data$pvalue < 0.01, "Downregulated", "Not Significant")
)

# Summary counts
table(rna_data$Significant)

```


### Volcano Plot  
Figure 1: Volcano plot showing significantly upregulated (red) and downregulated (blue) genes.
```{r}
library(ggplot2)

ggplot(rna_data, aes(x = log2FoldChange, y = -log10(pvalue), color = Significant)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Upregulated" = "red", 
                                "Downregulated" = "blue", 
                                "Not Significant" = "grey")) +
  theme_minimal() +
  labs(title = "Volcano Plot of RNA-seq Data",
       x = "log2 Fold Change",
       y = "-log10(p-value)")
```

## Top Differentially Expressed Genes
```{r}
# Upregulated genes
up_genes <- subset(rna_data, log2FoldChange > 1 & pvalue < 0.01)

# Downregulated genes
down_genes <- subset(rna_data, log2FoldChange < -1 & pvalue < 0.01)

# Top 5 upregulated
top5_up <- head(up_genes[order(up_genes$pvalue), ], 5)

# Top 5 downregulated
top5_down <- head(down_genes[order(down_genes$pvalue), ], 5)

```

```{r}
knitr::kable(top5_up, caption = "Top 5 Upregulated Genes")
knitr::kable(top5_down, caption = "Top 5 Downregulated Genes")

```

## Interpretation and Discussion

The volcano plot highlights a small set of genes with significant changes in expression.
A total of 19 genes were significantly upregulated and 91 genes were significantly downregulated.
Among the upregulated genes, EMILIN2 and DTHD1 showed the lowest p-values, suggesting strong statistical evidence of increased expression following treatment. Conversely, TBX5 was the most strongly downregulated gene (log2FC = –2.1), followed by LAMA2 and IFITM1, all of which may play roles in disease progression or treatment response.
These genes may represent key regulatory candidates involved in the biological condition being studied.

Further downstream analyses, such as Gene Ontology enrichment or pathway analysis, would help to link these genes to functional biological processes.
Validation using independent methods (e.g., qPCR) is recommended to confirm these expression changes.

## Conclusion
This analysis identified a small number of genes with significant expression changes between the two conditions. The top candidates (e.g., *TBX5*, *EMILIN2*) may play important roles in the underlying biological process and warrant further investigation.



